Measurement Clusterization in Bayesian D-optimal Designs in Infinite Dimensions

### 2024-07-31 latex/prelims.tex
prelims finished

\section{Preliminaries and Notation}\label{section:prelim}
Here we present the basics of Bayesian inverse problems over hilbert
spaces. Since we are ultimately interested in inferring a function
over some domain, we will keep in mind that these Hilbert spaces
should really be thought of as function spaces. A deeper treatment of
the foundations for inverse problems over function spaces can be found
in \cite{Stuart10}.

@@

The connection between parameter and observation spaces is made by the
\emph{forward operator} \(\fwd: \hilp \to \hilo\). We assume the
forward operator \(\fwd\) is linear. In the inverse problem of the 1D
heat equatin, the forward operator is determined by the time evolution
of the 1D heat equation \eqref{eq:heat1} and \eqref{eq:heat2}.

Measurements are taken via a \emph{measurement operator} \(\obs\). It
is common for the measurement and forward operators to be merged
\(\tmp := \obs \fwd\) \cite{AlexanderianGloorGhattas14}, but the
analysis carried out in the following sections requires that \(\fwd\)
and \(\obs\) are explicitly separated (as in
\cite{attia2022stochastic, cvetkovic2023choosing}). Note that \(\obs
\in ( \hilo^* )^m\), where \(m\) is the number of measurements
taken\footnote{The alert reader will likely ask how do we reconcile
point measurements $\delta_x$ as suggested by the formulation of the
1D heat equation with workin in Hilbert spaces. We don't. This entire
study could be formulated for Banach spaces where point measurements
are allowed. We avoid this technical difficulty because: (1) it does
not add much to the study, (2) it is beyond the technical abilities of
the author}. Entries \(\meas_j, j=1,\dots,m\) of the measurement
operator \(\obs\) are called \emph{measurements}:

\begin{equation*}%\label{eq:O}
  \obs u = (\meas_1(u), \dots, \meas_m(u) )^t \in \R^m,\ u \in \hilo.
\end{equation*}

Data is acquired via noisy observations, and we consider two types of
error terms: Spatially correlated model error \(\eps' \sim
\normal(0,\modcov)\) with \(\modcov\) a covariance operator; and
observation error denoted \(\eps \sim \normal(0, \sigma^2 I_m)\), with
\(I_m \in \mathbb{R}^{m \times m}\) the identity. Both error terms and
the prior (see below) are assumed independent of each other. Thus,
data is acquired via
\begin{align}\label{eq:inverse_problem}
  \data := \obs (\fwd \param + \eps') + \eps = \obs \fwd \param + \obs \eps' + \eps.
\end{align}

@@

\subsection{Bayesian Linear Inverse Problems}\label{subsec:bayesian_inverse_problems}
In the previous section, we saw how a parameter $u\in \hilp$ is
transported to the observation space via the forward operaror $\fwd u
\in \hilo$, how observations are generated from a parameter via $\obs
\fwd u$ and how observations and noise give rise to data $\data$. It
is time to formulate the process of inferring the parameter as a
Bayesian inverse problem. The first ingredient we need is a prior.

We take a Gaussian prior \(\param \sim \pr = \normal(\prmean
,\prcov)\) with some appropriate covariance operator \(\prcov\) on
\(\hilp\) \cite{Stuart10}. For example, for the inverse problem of the
1D heat equation we chose $\prcov = (-\Delta)^{-1}$ --- see Section
\ref{subsec:toy}. Note that \(\fwd \prcov \fwd^*\) is the prior
covariance in \(\hilo\) \cite{Stuart10}, and as such is assumed
invertible --- an assumption which we will use later (if \(\fwd\) has
a nontrivial kernel we utilize Occam's Razor and ignore said kernel
altogether).

Since $\fwd$ is linear and $\pr$ is Gaussian --- the posterior
\(\post\) is Gaussian as well. We do not utilize the posterior mean in
this study, but the posterior covariance operator $\postcov$ is given
by the known formula\cite{Stuart10}:
\begin{align}\label{eq:postcov}
  \postcov = (\prcov^{-1} + \fwd^* \obs^* \Sigma^{-1} \obs \fwd
  )^{-1}.
\end{align}

\subsection{Bayesian D-Optimal Designs}\label{subsec:D_optimal_design} 
A Bayesian D-optimal design maximizes the expected KL divergence
between posterior \(\post\) and prior measures \(\pr\). For arbitrary
posterior and prior measures, the KL divergence is defined,
analogously to eq.~\ref{eq:basic_KL}, via the Radon-Nikodym
derivative:
\begin{equation*}
  D_{KL}(\post||\pr) = \int \log \frac{\der \post}{\der \pr}(\param) \der \post(\param).
\end{equation*}

@@

Note that in \cite{AlexanderianGloorGhattas14,
  alexanderian2018efficient}, results are stated for \(\Sigma=I\)
(implied by \(\modcov = 0,\sigma^2=1\)), but these results also hold
for more general covariance matrices
\cite[p. 681]{AlexanderianGloorGhattas14}.

>>>>

\section{Preliminaries and Notation}\label{section:prelim}
Here we present the basics of Bayesian inverse problems over Hilbert
spaces. Since we are ultimately interested in inferring a function
over some domain, we will keep in mind that these Hilbert spaces
should really be thought of as function spaces. A deeper treatment of
the foundations for inverse problems over function spaces can be found
in \cite{Stuart10}.

@@

The connection between parameter and observation spaces is made by the
\emph{forward operator} \(\fwd: \hilp \to \hilo\). We assume the
forward operator \(\fwd\) is linear. In the inverse problem of the 1D
heat equation, the forward operator is determined by the time
evolution of the 1D heat equation \eqref{eq:heat1} and
\eqref{eq:heat2}, so $u(\cdot, T) = \fwd u_0$.

Measurements are taken via a linear \emph{measurement operator}
\(\obs\), which is the concatenation of linear functionals
$\meas_1,\dots,\meas_m$ we call \emph{measurements}:
\begin{equation*}%\label{eq:O}
  \obs u = (\meas_1(u), \dots, \meas_m(u) )^t \in \R^m,\ u \in \hilo.
\end{equation*}
%% \footnote{It is common for the measurement and forward
%% operators to be merged \(\tmp := \obs \fwd\)
%% \cite{AlexanderianGloorGhattas14}, but the analysis carried out in the
%% following sections requires that \(\fwd\) and \(\obs\) are explicitly
%% separated, as in \cite{attia2022stochastic,
%%   cvetkovic2023choosing}.}
Thus, \(\obs \in ( \hilo^* )^m\), where \(m\) is the number of
measurements taken\footnote{The alert reader will likely ask how do we
reconcile point measurements $\delta_x$ as suggested by the
formulation of the 1D heat equation with working in Hilbert spaces. We
don't. We follow standard practice in the literature and restrict our
analysis to Hilbert spaces. We can satisfy ourselves with the fact
that point evaluations could be approximated in a standard Hilbert
space like $L^2(\Omega)$.}.

Data is acquired via noisy observations, and we consider two types of
error terms: Spatially correlated model error \(\eps' \sim
\normal(0,\modcov)\) with \(\modcov\) a covariance operator; and
observation error denoted \(\eps \sim \normal(0, \sigma^2 I_m)\), with
\(I_m \in \mathbb{R}^{m \times m}\) the identity. Both error terms and
the prior (see Section \ref{subsec:bayesian_inverse_problems} below)
are assumed independent of each other. Thus, data is acquired via
\begin{align}\label{eq:inverse_problem}
  \data := \obs (\fwd \param + \eps') + \eps = \obs \fwd \param + \obs \eps' + \eps.
\end{align}

@@

\subsection{Bayesian Linear Inverse Problems}\label{subsec:bayesian_inverse_problems}
In the previous section, we saw how a parameter $u\in \hilp$ is
transported to the observation space via the forward operator $\fwd u
\in \hilo$, how observations are generated from a parameter via $\obs
\fwd u$ and how observations and noise give rise to data $\data$. It
is time to formulate the process of inferring the parameter as a
Bayesian inverse problem. We have already defined the likelihood in
the previous section, and now we will define the prior.

We take a Gaussian prior \(\param \sim \pr = \normal(\prmean
,\prcov)\) with some appropriate covariance operator \(\prcov\) on
\(\hilp\) \cite{Stuart10}. For example, for the inverse problem of the
1D heat equation we chose $\prcov = (-\Delta)^{-1}$, as described in
Section \ref{subsec:toy}. Note that \(\fwd \prcov \fwd^*\) is the
prior covariance in \(\hilo\) \cite{Stuart10}, and as such is assumed
invertible --- an assumption which we will use later (if \(\fwd\) has
a nontrivial kernel we utilize Occam's Razor and ignore said kernel
altogether).

Since $\fwd$ is linear and $\pr$ is Gaussian --- the posterior
\(\post\) is Gaussian as well. We do not utilize the posterior mean in
this study, but the posterior covariance operator $\postcov$ is given
by the known formula \cite{Stuart10}:
\begin{align}\label{eq:postcov}
  \postcov = (\prcov^{-1} + \fwd^* \obs^* \Sigma^{-1} \obs \fwd
  )^{-1}.
\end{align}

\subsection{Bayesian D-Optimal Designs}\label{subsec:D_optimal_design} 
A Bayesian D-optimal design maximizes the expected KL divergence
between posterior \(\post\) and prior measures \(\pr\). For arbitrary
posterior and prior measures, the KL divergence is defined,
analogously to eq.~\eqref{eq:basic_KL}, via the Radon-Nikodym
derivative:
\begin{equation*}
  D_{KL}(\post||\pr) = \int \log \frac{\der \post}{\der \pr}(\param) \der \post(\param).
\end{equation*}

@@

In \cite{AlexanderianGloorGhattas14, alexanderian2018efficient},
results are stated for \(\Sigma=I\) (implied by \(\modcov =
0,\sigma^2=1\)), but these results also hold for more general
covariance matrices \cite[p. 681]{AlexanderianGloorGhattas14}.
